Forecasting the 2024 Election: Trump Dominates the South, Harris Leads in the Midwest

### 2024-10-21 paper/paper.qmd
committing rendered paper.qmd

## Diagnostics
@tbl-modelsummary presents the results of the logistic regression model predicting the winner of the 2024 US Presidential Election in each state based on polling percentages for Donald Trump and Kamala Harris. The coefficients indicate that an increase in Trump’s polling percentage (\( \beta = 8.72 \)) is positively associated with the probability of winning a state, whereas an increase in Harris’s polling percentage (\( \beta = -31.63 \)) is negatively associated with Trump’s probability of winning. The standard errors for both predictors are quite large, suggesting high variability in the estimates and possible uncertainty in the model.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelsummary
#| tbl-cap: "Explanatory models of election outcome based on polling percentages for Trump and Harris"
#| warning: false

# Ensure the model exists
if (exists("logistic_model")) {
  # Create and display a model summary table in LaTeX format for PDF
  modelsummary::modelsummary(
    list("First Model" = logistic_model),
    statistic = c("std.error", "p.value"),  # Use standard error and p-value
    fmt = 2,  # Format numbers to 2 decimal places
    output = "latex"  # Use LaTeX for PDF output
  )
} else {
  print("Model not found")
}

@@

The model diagnostics, including the AIC (6.0) and BIC (8.1), suggest that the model fits the data reasonably well given the small sample size of 15 observations. However, the low log-likelihood (0.000) and root mean squared error (RMSE = 0.00) indicate that the model may not be well-calibrated, likely due to the small dataset or the structure of the data. These results emphasize the importance of using a larger sample size.

>>>>

## Diagnostics

Out-of-sample testing was conducted to test our model’s reliability in predicting potential presidential election outcomes. We train our model on a training set of our data and then test its performance on the test set of our data, we evaluate how well the model generalizes to new data that it was not trained on. This process helps reveal whether the model is overfitting to the training data or truly capturing the underlying patterns that apply more broadly. The test produces an accuracy score, RMSE, and a confusion matrix, each measuring different aspects of the model’s prediction. Accuracy is the proportion of correctly predicted outcomes compared to the total number of predictions made in the test set.

```{r}

#| echo: false
#| message: false
#| warning: false

# Out-of-Sample Testing of Our Model
# Load necessary libraries
library(caret)
library(rstanarm)
library(ggplot2)
library(dplyr)

# Split the data into training (70%) and test (30%) sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(state_average$winner, p = 0.7, list = FALSE)
train_data <- state_average[trainIndex, ]
test_data <- state_average[-trainIndex, ]

# Train the Bayesian logistic regression model on the training data
logistic_model_train <- stan_glm(winner ~ Trump_pct + Harris_pct, data = train_data, 
                                 family = binomial, prior = student_t(df = 7), 
                                 chains = 4, iter = 2000, refresh = 0, verbose = FALSE)

# Make predictions on the test set
test_predictions <- posterior_predict(logistic_model_train, newdata = test_data, draws = 1)

# Convert predictions to binary outcome (1 if Trump wins, 0 if Harris wins)
test_pred_binary <- ifelse(colMeans(test_predictions) > 0.5, 1, 0)

# Calculate accuracy
accuracy <- mean(test_pred_binary == test_data$winner)
accuracy_message <- paste("Test Accuracy: ", round(accuracy * 100, 2), "%")
print(accuracy_message)

# Optionally, calculate RMSE
rmse_value <- sqrt(mean((test_pred_binary - test_data$winner)^2))
rmse_message <- paste("Test RMSE: ", round(rmse_value, 4))
print(rmse_message)

# Confusion Matrix for additional diagnostics
conf_matrix <- table(Predicted = test_pred_binary, Actual = test_data$winner)
print(conf_matrix)

@@

Our accuracy test generated a 75% score for the model, which means that our model is quite reliable for making predictions about state-level election outcomes. Since presidential elections hinge on state-level wins, a reliable state-by-state prediction would imply a reliable national outcome prediction. The RMSE value generated for our model is 0.5, which indicates that the model-predicted probabilities are close to the true outcomes. The confusion matrix breaks down the model’s predictions into four categories: 1. correctly predicted Trump wins, 2. correctly predicted Harris wins, 3. Incorrectly predicted Trump wins, 4. Incorrectly predicted Harris wins. The generated confusion matrix for our model all had values of 0 to 2 for all four categories, indicating a balanced prediction between the two candidates, and suggesting the reliability of our model.
